HEP Software Foundation Community White Paper Working Group -- Detector Simulation

### 2018-03-15 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex

\begin{itemize}
\item
  reviewing the physics models assumptions, approximations and
  limitations in order to achieve higher precision, and to extend the
  validity of models up to energies of the order of 100 TeV foreseen
  with the Future Circular Collider (FCC) project;
\item
  redesigning, developing, commissioning detector simulation toolkits to
  be more efficient when executed on current vector CPUs and emerging
  new architectures, including GPGPUs where use of SIMD vectorisation is
  vital; this includes porting and optimising the experiments'
  simulation applications to allow exploitation of large HPC facilities;
\item
  exploring different Fast Simulation options, where the full detector
  simulation is replaced, in whole or in part, by computationally
  efficient techniques; an area of investigation is common frameworks
  for fast tuning and validation;
\item
  developing, improving, and optimising geometry tools that can be
  shared among experiments to make the modeling of complex detectors
  computationally more efficient, modular, and transparent;
\item
  developing techniques for background modeling, including contributions
  of multiple hard interactions overlapping the event of interest in
  collider experiments (pile-up);
\item
  revisiting digitisation algorithms to improve performance and
  exploring opportunities for code sharing among experiments;
\item
  recruiting, training, retaining, human resources in all areas of
  expertise pertaining to the simulation domain, including software and
  physics.
\end{itemize}

The complexity and diversity of future experiments results in an
enormous demand for CPU, and detailed detector simulation already
consumes the plurality of CPU resources of most current HEP experiments.
Some progress has already been made in improving software performance,
for example by adding multithreading capability to Geant4, and by
redesigning the geometry library into the VecGeom package that exploits
vectorisation. These R\&D programmes need to continue in order to
benefit from the various forms of parallelism offered by modern
architectures. For example, the goal of the GeantV R\&D project is to
build a realistic prototype application for a complex experiment with
the aim of demonstrating a big improvement in CPU time. The main
architectural feature of this prototype is the implementation of
particle-level, i.e. track-level, parallelisation for more efficient use
of SIMD vectorisation and data locality. Whenever possible, new
developments resulting from the R\&D programme will be tested in
realistic prototypes and then be integrated, validated, and deployed in
a timely fashion in Geant4 in order to maximise the benefits to
experiments at the earliest opportunity.

@@

Muon experiments such as Muon g-2 and Mu2e also face large simulation
challenges, being Geant4 the primary simulation toolkit for
all stages of these experiments. Because they are searching for extremely rare effects, they
must grapple with very low signal to background ratios, and the modeling
of low cross-section background processes. Often the physics of these
processes is not well understood, introducing large systematic
uncertainties. Additionally, the size of the computational problem is a
serious challenge, as large simulation runs are required to adequately
sample all relevant areas of experimental phase space, even when
techniques to minimise the required computations are used. Another
aspect related to the background events is the need to simulate the
effects of low energy neutrons, which requires large computational
resources. For example, Mu2e used a total of approximately 60 million
CPU hours, or 83 thousand core months, in a 2016 simulation campaign to
produce the Full Simulation sample needed to complete a technical design
report delivered in 2016.

>>>>

\begin{itemize}
\item
  reviewing the physics models' assumptions, approximations and
  limitations in order to achieve higher precision, and to extend the
  validity of models up to energies of the order of 100 TeV foreseen
  with the Future Circular Collider (FCC) project;
\item
  redesigning, developing, commissioning detector simulation toolkits to
  be more efficient when executed on current vector CPUs and emerging
  new architectures, including GPGPUs where use of SIMD vectorisation is
  vital; this includes porting and optimising the experiments'
  simulation applications to allow exploitation of large HPC facilities;
\item
  exploring different Fast Simulation options, where the full detector
  simulation is replaced, in whole or in part, by computationally
  efficient techniques; an area of investigation is common frameworks
  for fast tuning and validation;
\item
  developing, improving, and optimising geometry tools that can be
  shared among experiments to make the modeling of complex detectors
  computationally more efficient, modular, and transparent;
\item
  developing techniques for background modeling, including contributions
  of multiple hard interactions overlapping the event of interest in
  collider experiments (pile-up);
\item
  revisiting digitisation algorithms to improve performance and
  exploring opportunities for code sharing among experiments;
\item
  recruiting, training, retaining, human resources in all areas of
  expertise pertaining to the simulation domain, including software and
  physics.
\end{itemize}

The complexity and diversity of future experiments results in an
enormous demand for CPU, and detailed detector simulation already
consumes a large fraction of CPU resources of most current HEP experiments.
Some progress has already been made in improving software performance,
for example by adding multithreading capability to Geant4, and by
redesigning the geometry library into the VecGeom package that exploits
vectorisation. These R\&D programmes need to continue in order to
benefit from the various forms of parallelism offered by modern
architectures. For example, the goal of the GeantV R\&D project is to
build a realistic prototype application for a complex experiment with
the aim of demonstrating a big improvement in CPU time. The main
architectural feature of this prototype is the implementation of
particle-level, i.e. track-level, parallelisation for more efficient use
of SIMD vectorisation and data locality. Whenever possible, new
developments resulting from the R\&D programme will be tested in
realistic prototypes and then be integrated, validated, and deployed in
a timely fashion in Geant4 in order to maximise the benefits to
experiments at the earliest opportunity.

@@

Muon experiments such as Muon g-2 and Mu2e also face large simulation
challenges, as Geant4 is the primary simulation toolkit for
all stages of these experiments. Because they are searching for extremely rare effects, they
must grapple with very low signal to background ratios, and the modeling
of low cross-section background processes. Often the physics of these
processes is not well understood, introducing large systematic
uncertainties. Additionally, the size of the computational problem is a
serious challenge, as large simulation runs are required to adequately
sample all relevant areas of experimental phase space, even when
techniques to minimise the required computations are used. Another
aspect related to the background events is the need to simulate the
effects of low energy neutrons, which requires large computational
resources. For example, Mu2e used a total of approximately 60 million
CPU hours, or 83 thousand core months, in a 2016 simulation campaign to
produce the Full Simulation sample needed to complete a technical design
report delivered in 2016.
